---
title: "Stats 100: Project - MLB"
author: "Niall Kehoe"
date: "05/11/2023"
header-includes:
   - \usepackage{bbm, amsmath,amsfonts,amsthm,amssymb,mathrsfs,amsxtra,amscd,latexsym, xcolor, graphicx, fvextra}
   - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
output: pdf_document
---

\newcommand{\Var}{\mathrm{Var}}

```{r}
suppressPackageStartupMessages(library(knitr)) #makes pdfs
suppressPackageStartupMessages(library(latex2exp))
suppressPackageStartupMessages(library(ggplot2)) #makes nice plots
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(glmnet))
suppressPackageStartupMessages(library(predtools))
suppressPackageStartupMessages(library(psych))
knitr::opts_chunk$set(echo = TRUE)

# Checks if package is installed, installs if necessary, and loads package for current session

pacman::p_load(
  lubridate,  # general package for handling and converting dates  
  parsedate,   # has function to "guess" messy dates
  aweek,      # another option for converting dates to weeks, and weeks to dates
  zoo,        # additional date/time functions
  tidyverse,  # data management and visualization  
  rio)        # data import/export
```

# Acknowledgements

NO ACKNOWLEDGEMENTS

# Bradley-Terry Ratings

```{r}

#df_final = read.csv("NBA_2022_23_3pdata_final.csv")
mlb2022 = read.csv("mlb2022.csv") #%>% filter(Date > )

#Removing last row MLB as it is totals
mlb2022 <- slice(mlb2022, 1:(n() - 2)) 

View(mlb2022)


```

```{r}


```

```{r}

```

```{r}

```

```{r}

#length(names(sort(bt_normal$coefficients[-1], decreasing=TRUE)))
teamNames = names(sort(bt_normal$coefficients[-1], decreasing=TRUE))
append(teamNames, "Washington.Wizards")
# DONT FORGET TO ADD BACK IN BASELINE at 0

NBASalaries = read.csv("basketball-2021-salary.csv")#  %>% filter(Team %in% goodHosp)
#View(NBASalaries)

NBASalaries
#NBASalaryDict = {}

#for(i in 1:nrow(NBASalaries)) {
#  formattedName = NBASalaries[i,]$Team
#  formattedName <- str_replace_all(formattedName, " ", ".")
#  NBASalaryDict[formattedName] = substring(NBASalaries[i,]$X2022.23,2)
#}

#NBASalaryDict
```

```{r}

coeffs = bt_normal$coefficients[-1]
coeffs["Washington.Wizards"] = 0

NBASalaries <- NBASalaries[order(NBASalaries$Team,decreasing=FALSE),]

dfResult <- data.frame(coeffs, NBASalaries$X2021.22, NBASalaries$Rk)# %>% mutate(Salaries = NBASalaries$X2022.23 %>% str_remove_all("\\$,") %>% as.numeric())
dfResult$Salaries = as.numeric(gsub("\\$", "", dfResult$NBASalaries.X2021.22))
dfResult$Salaries = dfResult$Salaries / 1000000

View(dfResult)

# Plot the chart for cars with weight between 2.5 to 5 and mileage between 15 and 30.
plot(x = dfResult$Salaries, y = dfResult$coeffs,
   xlab = "Salaries (millions)",
   ylab = "Performance (pt diff vs Washington Wizards)",	 
   main = "Salaries vs Performance"
)

modelCoeff = lm(coeffs ~ Salaries, data = dfResult)
summary(modelCoeff)


# Divide by average salary -try 
```

```{r}

hist(dfResult$Salaries, col = 'skyblue3', breaks = 15)
```

```{r}

dfResult <- dfResult[order(dfResult$coeffs,decreasing=FALSE),]
ranksMap = seq(1, 30, by=1)
dfResult$coeffRank = ranksMap
dfResult
```

```{r}


# Plot the chart for cars with weight between 2.5 to 5 and mileage between 15 and 30.
plot(x = dfResult$Salaries, y = dfResult$coeffRank,
   xlab = "Salaries (millions)",
   ylab = "Performance by ranking in BT",	 
   main = "Salaries vs Performance"
)

cor(dfResult$Salaries, dfResult$coeffRank)

modelRank = lm(coeffRank ~ Salaries, data = dfResult)
summary(modelRank)

# Log change

plot(x = log(dfResult$Salaries), y = dfResult$coeffRank,
   xlab = "Salaries (millions)",
   ylab = "Performance by ranking in BT",	 
   main = "log(Salaries) vs Performance"
)
modelRank = lm(coeffRank ~ log(Salaries), data = dfResult)
summary(modelRank)
```

```{r}

# NBA RANKINGS W-L

standings = read.csv("basketball-2021-standings.csv") #%>% filter(Date > )
dfStandings <- data.frame(standings$Team, ranksMap) #NBASalaries$X2021.22, NBASalaries$Rk)

NBASalaries <- NBASalaries[order(NBASalaries$Team,decreasing=FALSE),]
dfStandings <- dfStandings[order(dfStandings$standings.Team,decreasing=FALSE),]

dfStandings$Salaries = NBASalaries$X2021.22
dfStandings$Salaries = as.numeric(gsub("\\$", "", dfStandings$Salaries))
dfStandings$Salaries = dfStandings$Salaries / 1000000

#dfStandings <- dfStandings[order(dfStandings$dfStandings.Team,decreasing=FALSE),]


colnames(dfStandings) <- c("Team", "ranking", "Salaries")


dfStandings$rankingA = 30- dfStandings$ranking

View(dfStandings)

plot(x = dfStandings$Salaries, y = dfStandings$ranking,
   xlab = "Salaries (millions)",
   ylab = "Performance by ranking No. wins (lower better)",	 
   main = "Salaries vs Performance"
)

plot(x = dfStandings$Salaries, y = dfStandings$rankingA,
   xlab = "Salaries (millions)",
   ylab = "Performance by rankingA (higher better)",	 
   main = "Salaries vs Performance"
)


modelRank = lm(rankingA ~ Salaries, data = dfStandings)
summary(modelRank)

# Try
# Total score / $ spent
```

## 2.4

In either model you fit, what are some model assumptions that you find problematic that might lead to shortcomings in the results? [\textbf{2 points}]

Shortcomings: styles of play affecting which teams beat which, injuries affecting performance, small number of games in training set, trading of players mid season.
